---
title: "describe bivariate for Kaggle challenge: Tabular Playground Series - Feb 2022"
author: "Sascha Siegmund"
date: "`r Sys.Date()`"
output: 
  github_document:
    fig_width: 10
    fig_height: 7
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  fig.path = "nb_figs/multi_",
  warning = FALSE,
  message = FALSE
)
```



## purpose of notebook

  - [ ] describe & visualize relationship between variables (multivariate)
  - [ ] gather interesting observations for further investigation
  - [ ] gather possible new features for extraction
  
  
## insights 

  - 


## load packages

```{r load packages, setup, message=FALSE}
library(tidyverse) # tidy data frame
library(plotly) # make ggplots interactive
library(ggmosaic) # for mosaic plots 
```


## import data

```{r, message=FALSE}
df <- read_csv('../data/train.csv') 
```


## overview

```{r}
head(df)
```
```{r}
summary(df)
```


## overview of all bivariate combinations

- scagnostics is used to produce all the plot combinations, but is also capable of calculating a lot of different measures to classify the bivariate plot 

```{r}
# Scagnostics (scatterplot diagnostics) summarize potentially interesting patterns in 2d scatterplots
name = c('')
tmp_df <- df %>% 
  # select() %>%  # remove categorical and time
  na.omit() 

library(scagnostics)
# scagnostics_results[,1]
# scagnostics_results[1,]

scagnostics_results <- scagnostics(tmp_df)
head(scagnostics_results)

e <- scagnosticsExemplars(scagnostics_results)
o <- scagnosticsOutliers(scagnostics_results)
g <- scagnosticsGrid(scagnostics_results)

# look at outliers
# o[o]
# go <- g[o,]
# par(mfrow = c(1,1))
# plot(tmp_df[[go$x]], tmp_df[[go$y]], pch=19,
#      xlab=names(tmp_df)[go$x], ylab=names(tmp_df)[go$y])

# find and plot exemplars
e[e]
ge <- g[e,]
par(mfrow = c(1,1))
for(i in 1:dim(ge)[1])
  plot(tmp_df[[ge$x[i]]], tmp_df[[ge$y[i]]], pch=19,
       xlab=names(tmp_df)[ge$x[i]], ylab=names(tmp_df)[ge$y[i]])

par(mfrow = c(2,2))
for(i in 1:dim(g)[1])
  plot(tmp_df[[g$x[i]]], tmp_df[[g$y[i]]], pch=19,
       xlab=names(tmp_df)[g$x[i]], ylab=names(tmp_df)[g$y[i]])
```


## explore bivariate numeric over numeric, many observations

- 

```{r}
# two variables, continuous x, continuous y, show trend and distribution
name = c('A4T2G4C0', 'A3T4G3C0')
tmp_df <- df %>% rename(x = name[1], y = name[2]) %>% select(x, y)

p1 <- tmp_df %>%
  ggplot(aes(x = x, y = y)) +
    stat_density_2d(aes(fill = after_stat(density)), 
                    geom = "raster", contour = FALSE, show.legend = FALSE) +
    scale_fill_viridis_c() +
    # geom_quantile(alpha = 0.7, color = 'darkorange') +
    geom_smooth() +
    theme_minimal() +
    ggtitle(paste("trend of", name[2], "over", name[1], sep=" "))
p1 <- ggplotly(p1) %>% layout(xaxis = list(title = name[1]), yaxis = list(title = name[2]))

p2 <- tmp_df %>%
  ggplot(aes(x = x)) +
    stat_density(geom="line") + 
    theme_minimal()
p2 <- ggplotly(p2) %>% layout(yaxis = list(showticklabels = FALSE, showgrid = FALSE, title = ''), 
                              xaxis = list(showticklabels = FALSE, showgrid = FALSE, title = ''))

p3 <- tmp_df %>%
  ggplot(aes(x = 1, y = x)) +
    geom_boxplot() +
    theme_minimal() +
    coord_flip() 
p3 <- ggplotly(p3) %>% layout(yaxis = list(showticklabels = FALSE, showgrid = FALSE, title = ''), 
                              xaxis = list(showticklabels = FALSE, showgrid = FALSE, title = ''))

p4 <- tmp_df %>%
  ggplot(aes(x = y)) +
    stat_density(geom="line") + 
    coord_flip() +
    theme_minimal()
p4 <- ggplotly(p4) %>% layout(yaxis = list(showticklabels = FALSE, showgrid = FALSE, title = ''), 
                              xaxis = list(showticklabels = FALSE, showgrid = FALSE, title = ''))

p5 <- tmp_df %>%
  ggplot(aes(x = 1, y = y)) +
    geom_boxplot() +
    theme_minimal() 
p5 <- ggplotly(p5) %>% layout(yaxis = list(showticklabels = FALSE, showgrid = FALSE, title = ''), 
                              xaxis = list(showticklabels = FALSE, showgrid = FALSE, title = ''))

# merge figures into one plot, via subplots, https://plotly-r.com/arranging-views.html
fig <- subplot(p2, plotly_empty(), plotly_empty(),
               p3, plotly_empty(), plotly_empty(),
               p1, p5, p4, 
               nrows = 3, margin = 0, heights = c(0.1, 0.035, 0.865), widths = c(0.92, 0.02, 0.06), 
               shareX = TRUE, shareY = TRUE, titleX = TRUE, titleY = TRUE) %>% layout()

fig
```


## explore bivariate numeric over categorical

- 

```{r}
# two variables, continuous/discrete y, categorical x, show trend and distribution
name = c('target', 'A3T4G3C0')
tmp_df <- df %>% rename(c = name[1], y = name[2]) %>% select(c, y) %>% mutate(c = fct_infreq(factor(c))) %>% 
  add_count(c, y)

so.q2 <- function(y){ quantile(y, na.rm = TRUE)[2] }
so.q4 <- function(y){ quantile(y, na.rm = TRUE)[4] }

p1 <- tmp_df %>%
  ggplot(aes(x = c, y = y)) +
    geom_violin(aes(color = c), scale = "area", fill = NA) +
    geom_boxplot(fill = NA, lwd = 0.4, outlier.alpha = 0.3, width =0.1) +
    geom_hline(aes(yintercept = quantile(tmp_df$y, na.rm = TRUE)[2]), color = 'darkorange', alpha = 0.5, lty=2) + 
    geom_hline(aes(yintercept = quantile(tmp_df$y, na.rm = TRUE)[3]), color = 'darkred', alpha = 0.5, lty=2) +
    geom_hline(aes(yintercept = quantile(tmp_df$y, na.rm = TRUE)[4]), color = 'darkorange', alpha = 0.5, lty=2) +
    stat_summary(aes(x = as.numeric(c)), fun = so.q2, geom = 'line', size = 0.5, color = 'darkorange') + 
    stat_summary(aes(x = as.numeric(c)), fun = 'median', geom = 'line', size = 0.5, color = 'darkred') +
    stat_summary(aes(x = as.numeric(c)), fun = so.q4, geom = 'line', size = 0.5, color = 'darkorange') + 
    geom_spoke(aes(x = as.numeric(c), radius = n/max(tmp_df$n)/2, angle = 2*pi, color = c),
               alpha = I(0.5), stat = "unique") + # x-offset= +n/2, radius= n, angle= pi for two-sided spoke plot
    scale_color_brewer(guide = "none", palette = "Set2") +
    coord_flip() +
    theme_minimal() +
    ggtitle(paste("distribution of", name[2], "grouped by", name[1], sep=" ")) 
p1 <- ggplotly(p1) %>% layout(xaxis = list(title = name[2]), yaxis = list(title = name[1]))

p2 <- tmp_df %>%
  ggplot(aes(x = y)) +
    stat_density(geom="line") + 
    theme_minimal()
p2 <- ggplotly(p2) %>% layout(yaxis = list(showticklabels = FALSE, showgrid = FALSE, title = ''), 
                              xaxis = list(showticklabels = FALSE, showgrid = FALSE, title = ''))

p3 <- tmp_df %>%
  ggplot(aes(x = 1, y = y)) +
    geom_boxplot() +
    theme_minimal() +
    coord_flip() 
p3 <- ggplotly(p3) %>% layout(yaxis = list(showticklabels = FALSE, showgrid = FALSE, title = ''), 
                              xaxis = list(showticklabels = FALSE, showgrid = FALSE, title = ''))

p5 <- tmp_df %>%
  ggplot() +
    geom_mosaic(aes(x=product(c), fill = c)) +
    geom_text(data = layer_data(ggplot2::last_plot(), 1) %>%  filter(.wt > 0),
               aes(x = (xmax + xmin)/2, y = (ymax + ymin)/2, 
                   label = paste0(.wt,"\n",scales::percent(.wt/nrow(tmp_df),  accuracy = 1))),
              size = 3.5) +
    scale_fill_brewer(guide = "none", palette = "Set2") +
    theme_minimal() +
    coord_flip()
p5 <- ggplotly(p5) %>% layout(yaxis = list(showticklabels = FALSE, showgrid = FALSE, title = ''), 
                              xaxis = list(showticklabels = FALSE, showgrid = FALSE, title = ''))

# merge figures into one plot, via subplots, https://plotly-r.com/arranging-views.html
fig <- subplot(p2, plotly_empty(),
               p3, plotly_empty(),
               p1, p5,
               nrows = 3, margin = 0, heights = c(0.07, 0.03, 0.9), widths = c(0.95, 0.05), 
               shareX = TRUE, titleX = TRUE, titleY = TRUE) %>% layout() %>% hide_legend()

fig
```


## explore bivariate numeric over categorical (many categories, density heatmap)

-

```{r}
name = c('A3T4G3C0', 'target')
tmp_df <- df %>% rename(x = name[1], c = name[2]) %>% select(x, c) %>% 
  mutate(c = factor(c))

p1 <- tmp_df %>% 
  ggplot(aes(x = x, y = c, group = c, fill = after_stat(density))) +
    geom_tile(stat = "density", show.legend = FALSE) +
    scale_fill_viridis_c() +
    theme_minimal() +
    ggtitle(paste("distribution of", name[2], "over", name[1], sep=" "))
p1 <- ggplotly(p1) %>% layout(xaxis = list(title = name[1]), yaxis = list(title = name[2]))

p2 <- tmp_df %>%
  ggplot(aes(x = x)) +
    stat_density(geom="line") + 
    theme_minimal()
p2 <- ggplotly(p2) %>% layout(yaxis = list(showticklabels = FALSE, showgrid = FALSE, title = ''), 
                              xaxis = list(showticklabels = FALSE, showgrid = FALSE, title = ''))

p3 <- tmp_df %>%
  ggplot(aes(x = 1, y = x)) +
    geom_boxplot() +
    theme_minimal() +
    coord_flip() 
p3 <- ggplotly(p3) %>% layout(yaxis = list(showticklabels = FALSE, showgrid = FALSE, title = ''), 
                              xaxis = list(showticklabels = FALSE, showgrid = FALSE, title = ''))

p5 <- tmp_df %>%
  ggplot(aes(x = c)) +
    geom_bar() +
    theme_minimal() +
    coord_flip()
p5 <- ggplotly(p5) %>% layout(yaxis = list(showticklabels = FALSE, showgrid = FALSE, title = ''), 
                              xaxis = list(showticklabels = FALSE, showgrid = FALSE, title = ''))

fig <- subplot(p2, plotly_empty(),
               p3, plotly_empty(),
               p1, p5,
               nrows = 3, margin = 0, heights = c(0.07, 0.03, 0.9), widths = c(0.95, 0.05), 
               shareX = TRUE, shareY = TRUE, titleX = TRUE, titleY = TRUE) %>% layout()

fig
```

```{r}
name = c('A4T2G4C0', 'target')
tmp_df <- df %>% rename(x = name[1], c = name[2]) %>% select(x, c) %>% 
  mutate(c = factor(c))

p1 <- tmp_df %>% 
  ggplot(aes(x = x, y = c, group = c, fill = after_stat(density))) +
    geom_tile(stat = "density", show.legend = FALSE) +
    scale_fill_viridis_c() +
    theme_minimal() +
    ggtitle(paste("distribution of", name[2], "over", name[1], sep=" "))
p1 <- ggplotly(p1) %>% layout(xaxis = list(title = name[1]), yaxis = list(title = name[2]))

p2 <- tmp_df %>%
  ggplot(aes(x = x)) +
    stat_density(geom="line") + 
    theme_minimal()
p2 <- ggplotly(p2) %>% layout(yaxis = list(showticklabels = FALSE, showgrid = FALSE, title = ''), 
                              xaxis = list(showticklabels = FALSE, showgrid = FALSE, title = ''))

p3 <- tmp_df %>%
  ggplot(aes(x = 1, y = x)) +
    geom_boxplot() +
    theme_minimal() +
    coord_flip() 
p3 <- ggplotly(p3) %>% layout(yaxis = list(showticklabels = FALSE, showgrid = FALSE, title = ''), 
                              xaxis = list(showticklabels = FALSE, showgrid = FALSE, title = ''))

p5 <- tmp_df %>%
  ggplot(aes(x = c)) +
    geom_bar() +
    theme_minimal() +
    coord_flip()
p5 <- ggplotly(p5) %>% layout(yaxis = list(showticklabels = FALSE, showgrid = FALSE, title = ''), 
                              xaxis = list(showticklabels = FALSE, showgrid = FALSE, title = ''))

fig <- subplot(p2, plotly_empty(),
               p3, plotly_empty(),
               p1, p5,
               nrows = 3, margin = 0, heights = c(0.07, 0.03, 0.9), widths = c(0.95, 0.05), 
               shareX = TRUE, shareY = TRUE, titleX = TRUE, titleY = TRUE) %>% layout()

fig
```




